GO enrichment

### 2022-02-25 analysis/GOenrichment.Rmd
add some more changes
Replace GO enrichment

```{r setup, include=FALSE}
library(topGO)
library(tidyverse)
library(rvest)
library(wesanderson)
library(httr)
library(stringi)
library(ggrepel)
library(eulerr)
library(UpSetR)
library(rrvgo)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

@@

# lost in seagrasses, each one uniquely vs all terrestrials
```{r eval=FALSE}
compare('lost_in_amphi_vs_all.txt', 'missing_amphi_vs_all_GO.txt', allgenes, annAT)
compare('lost_in_posi_vs_all.txt', 'missing_posi_vs_all_GO.txt', allgenes, annAT)
compare('lost_in_zmar_vs_all.txt', 'missing_zmar_vs_all_GO.txt', allgenes, annAT)
compare('lost_in_zmuel_vs_all.txt', 'missing_zmuel_vs_all_GO.txt', allgenes, annAT)
```

# lost in seagrasses vs all terrestrials
Here we compare GO terms for seagrasses and aquatics (seagrasses+duckweeds) vs all terrestrials

```{r eval=FALSE}

compare('lost_in_seagrasses.txt', 'missing_seagrasses_GO.txt', allgenes, annAT)
compare('lost_in_aquatics.txt', 'missing_aquatics_GO.txt', allgenes, annAT)
compare('only_in_seagrasses.txt', 'only_seagrasses_GO.txt', allgenes, annAT)

# Seagrass only comparisons

@@

load('annAtObject.RData')
sallgenes <- unique(unlist(sannAT))

compare('lost_in_amphi.txt', 'lost_in_amphi_vs_seagrasses_GO.txt', sallgenes, sannAT)
compare('lost_in_posi.txt', 'lost_in_posi_vs_seagrasses_GO.txt', sallgenes, sannAT)
compare('lost_in_zmar.txt', 'lost_in_zmar_vs_seagrasses_GO.txt', sallgenes, sannAT)
compare('lost_in_zmuel.txt', 'lost_in_zmuel_vs_seagrasses_GO.txt', sallgenes, sannAT)
```

@@

The opposite - which GO-terms are present only in one of the four species?

```{r eval=FALSE}
compare('only_in_amphi.txt', 'only_in_amphi_vs_seagrasses_GO.txt', sallgenes, sannAT)
compare('only_in_posi.txt', 'only_in_posi_vs_seagrasses_GO.txt', sallgenes, sannAT)
compare('only_in_zmar.txt', 'only_in_zmar_vs_seagrasses_GO.txt', sallgenes, sannAT)
compare('only_in_zmuel.txt', 'only_in_zmuel_vs_seagrasses_GO.txt', sallgenes, sannAT)
```

@@

for (f in list.files('./data/', pattern='GO.txt')){
  filename <- paste('./data/', f, sep='')
  go_and_pvalues <- readChar(filename, file.info(filename)$size)
  go_and_pvalues <- gsub(',', ' ', go_and_pvalues)

@@

  names(dat) <- c("term_ID","description","frequency", 'plot_X', 'plot_Y', 'log_size', 'value', 'uniqueness', 'dispensability', 'eliminated', 'representative')
  one.data <- dat
  if (one.data[1,] == '<html>') {next} # some datasets result in error
  one.data <- one.data [(one.data$plot_X != "null" & one.data$plot_Y != "null") & (! is.na(one.data$frequency)) & (one.data$value != 'null'), ];
  one.data <- type_convert(one.data)
  top_ten_values <- one.data %>% arrange(value) %>% head(10) %>% pull(term_ID) 
  one.data <- one.data %>% mutate(description2 = case_when(
    term_ID %in% top_ten_values ~ description, 
                                                           TRUE ~ ''))
  p1 <-
    ggplot(data = one.data, aes(plot_X, plot_Y, label = description2)) +
    geom_point(aes(colour = value, size = log_size), alpha = I(0.6)) +
    scale_size_area() +
    scale_colour_gradientn(colours = rev(wes_palette("Zissou1", 100, type = "continuous")), 
                           limits = c(min(one.data$value), 0)) +
    geom_point(
      aes(plot_X, plot_Y, size = log_size),
      shape = 21,
      fill = "transparent",
      colour = I (alpha ("black", 0.6))
    ) +
    scale_size_area() +
    geom_label_repel(max.overlaps = Inf, box.padding = 0.5,
                     aes(point.size = log_size)) +
    theme_minimal() +
    labs(y = "Semantic space x",
         x = "Semantic space y",
         color = 'Log(p-value)',
         size = 'Frequency') +
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank()
    ) +
    NULL
    #ggtitle(f)

@@

```{r results='asis'}
for( i in names(results_list)) {
  print(i)
  print(knitr::kable(results_list[[i]] %>% filter(Eliminated == FALSE) %>% select(TermID, Name, Value) %>% arrange(Value), caption=i))
  cat("\n")
}

@@

# Venn diagram GO overlaps

Let's also check which GO terms overlap between the 4 seagrasses!

@@

  cowplot::save_plot(file_name, plot_list[[i]], base_height=8)

@@

How many shared lost GO-terms are there? Hopefully, all four species will have lost the most GO-terms.
```{r}
a <- list(`P. australis` = results_list$missing_posi_vs_all_GO.txt$Name,
          `A. antarctica` = results_list$missing_amphi_vs_all_GO.txt$Name,
          `Z. marina` = results_list$missing_zmar_vs_all_GO.txt$Name,
          `Z. muelleri` = results_list$missing_zmuel_vs_all_GO.txt$Name,
          `A. thaliana` = results_list$missing_arabidopsis_vs_all_GO.txt$Name)

a_go_ids <- list(`P. australis` = results_list$missing_posi_vs_all_GO.txt$TermID,
          `A. antarctica` = results_list$missing_amphi_vs_all_GO.txt$TermID,
          `Z. marina` = results_list$missing_zmar_vs_all_GO.txt$TermID,
          `Z. muelleri` = results_list$missing_zmuel_vs_all_GO.txt$TermID,
          `A. thaliana` = results_list$missing_arabidopsis_vs_all_GO.txt$TermID)
```

```{r}
plot(euler(a), 
     quantities = TRUE,
     fill = rev(wes_palette("Zissou1", 15, type = 'continuous')),
    alpha = 0.5,
     labels = list(font = 4))
```

@@

What are the shared GO-terms in seagrasses, WITHOUT the Ara loss??
```{r}
setdiff(Reduce(intersect, a_no_ara), Reduce(intersect, a)) %>% enframe() %>% writexl::write_xlsx('./data/shared_lost_genes.xlsx')
```

@@

```{r}
writexl::write_xlsx(results_d, 'data/Lost_GO_terms_in_five_species.xlsx')
```

# after filtering for plant-specific GO-terms

@@

```{r}
results_d %>% filter(GOID %in% plantSpecificGO) %>% writexl::write_xlsx('data/Lost_GO_terms_in_five_species.PlantSpecific.xlsx')
```

@@

```{r}
plot(euler(newa), 
     quantities = TRUE,
     fill = rev(wes_palette("Zissou1", 15, type = 'continuous')),
    alpha = 0.5,
     labels = list(font = 4))
```

>>>>

```{r setup, include=FALSE}
library(topGO)
library(tidyverse)
library(rvest)
library(wesanderson)
library(httr)
library(stringi)
library(ggrepel)
library(eulerr)
library(UpSetR)
library(rrvgo)
library(kableExtra)
library(patchwork)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

@@

```
# Comparing the three big groups

```{r eval=FALSE}
# all genes shared - this runs out of memory
#compare('Lost_present_gene_lists/Genes_only_in_Aquatics_and_Seagrasses_and_Terrestrials.txt', 'output/GO_results_genes_in_Aquatics_and_Seagrasses_and_Terrestrials.csv', allgenes, annAT)

# Genes in aquatics and seagrasses - IMPORTANT This is NOT the union - this is the intersection!
# Union is below
compare('Lost_present_gene_lists/Genes_only_in_Aquatics_and_Seagrasses.txt', 'output/GO_results_genes_in_Aquatics_and_Seagrasses_NOT_Terrestrials.csv', allgenes, annAT)

# Genes only in aquatics
compare('Lost_present_gene_lists/Genes_only_in_Aquatics.txt', 'output/GO_results_genes_in_Aquatics_NOT_Seagrasses_NOT_Terrestrials.csv', allgenes, annAT)

# Genes only in seagrasses
compare('Lost_present_gene_lists/Genes_only_in_Seagrasses.txt', 'output/GO_results_genes_in_Seagrasses_NOT_Aquatics_NOT_Terrestrials.csv', allgenes, annAT)

# Genes only in terrestrials
compare('Lost_present_gene_lists/Genes_only_in_Terrestrials.txt', 'output/GO_results_genes_in_Terrestrials_NOT_Aquatics_NOT_Seagrasses.csv', allgenes, annAT)

# Genes union of Seagrass and Aquatics
compare('Lost_present_gene_lists/Genes_union_of_Seagrass_and_Aquatics_union.txt', 'output/GO_results_genes_union_Aquatics_and_Seagrasses_NOT_Terrestrials.csv', allgenes, annAT)

```
# Seagrass only comparisons

@@

load('sannAtObject.RData')
sallgenes <- unique(unlist(sannAT))

```
# lost in seagrasses

```{r eval=FALSE}
compare('Lost_present_gene_lists/Genes_lost_in_A_antarctica_not_other_seagrasses.txt', 'output/GO_results_genes_lost_A_antarctica_not_other_seagrasses.csv', sallgenes, sannAT)

compare('Lost_present_gene_lists/Genes_lost_in_Z_marina_not_other_seagrasses.txt', 'output/GO_results_genes_lost_Z_marina_not_other_seagrasses.csv', sallgenes, sannAT)

compare('Lost_present_gene_lists/Genes_lost_in_P_australis_not_other_seagrasses.txt', 'output/GO_results_genes_lost_P_australis_not_other_seagrasses.csv', sallgenes, sannAT)

compare('Lost_present_gene_lists/Genes_lost_in_Z_muelleri_not_other_seagrasses.txt', 'output/GO_results_genes_lost_Z_muelleri_not_other_seagrasses.csv', sallgenes, sannAT)
```

@@

```{r eval=FALSE}
compare('Lost_present_gene_lists/Genes_only_in_Z_marina_not_other_seagrasses.txt', 'output/GO_results_genes_only_Z_marina_not_other_seagrasses.csv', sallgenes, sannAT)

compare('Lost_present_gene_lists/Genes_only_in_Z_muelleri_not_other_seagrasses.txt', 'output/GO_results_genes_only_Z_muelleri_not_other_seagrasses.csv', sallgenes, sannAT)

compare('Lost_present_gene_lists/Genes_only_in_P_australis_not_other_seagrasses.txt', 'output/GO_results_genes_only_P_australis_not_other_seagrasses.csv', sallgenes, sannAT)

compare('Lost_present_gene_lists/Genes_only_in_A_antarctica_not_other_seagrasses.txt', 'output/GO_results_genes_only_A_antarctica_not_other_seagrasses.csv', sallgenes, sannAT)
```

@@

for (f in list.files('./output/', pattern='*.csv')){
  filename <- paste('./output/', f, sep='')
  go_and_pvalues <- readChar(filename, file.info(filename)$size)
  go_and_pvalues <- gsub(',', ' ', go_and_pvalues)

@@

  names(dat) <- c("term_ID","description","frequency", 'plot_X', 'plot_Y', 'log_size', 'value', 'uniqueness', 'dispensability', 'eliminated', 'representative')
  one.data <- dat
  if (one.data[1,] == '<html>') {next} # some datasets result in error
  one.data <- one.data [(one.data$plot_X != "null" & one.data$plot_Y != "null") & (! is.na(one.data$frequency)) & (one.data$value != 'null'), ];
  one.data <- suppressMessages(type_convert(one.data)) # guess data types, but also: shush
  top_ten_values <- one.data %>% arrange(value) %>% head(10) %>% pull(term_ID) 
  one.data <- one.data %>% mutate(description2 = case_when(
    term_ID %in% top_ten_values ~ description, 
    TRUE ~ ''))
  p1 <-
    ggplot(data = one.data, aes(plot_X, plot_Y, label = description2)) +
    geom_point(aes(colour = value, size = log_size), alpha = I(0.6)) +
    scale_size_area() +
    scale_colour_gradientn(colours = rev(wes_palette("Zissou1", 100, type = "continuous")), 
                           limits = c(min(one.data$value), 0)) +
    geom_point(
      aes(plot_X, plot_Y, size = log_size),
      shape = 21,
      fill = "transparent",
      colour = I (alpha ("black", 0.6))
    ) +
    scale_size_area() +
    geom_label_repel(max.overlaps = Inf, box.padding = 0.5,
                     aes(point.size = log_size), alpha=0.9) +
    theme_minimal() +
    labs(color = 'log(p-value)',
         size = 'Frequency') +
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      axis.title.x = element_blank(),
      axis.title.y = element_blank()
    ) +
    NULL
    #ggtitle(f)

@@

```{r results='asis'}
for( i in names(results_list)) {
  print(i)
  print(results_list[[i]] %>% filter(Eliminated == FALSE) %>% select(TermID, Name, Value) %>% arrange(Value) %>% kbl() %>%  kable_styling())
  cat("\n")
}

@@

@@

  cowplot::save_plot(file_name, plot_list[[i]], base_height=6)

@@

Let me also try and make some patchworks.

```{r fig.width=10}
# For some reason with this plot the legends aren't correctly collected by patchwork, 
# so I need to manually turn one off
patchy <- (plot_list[["GO_results_genes_in_Terrestrials_NOT_Aquatics_NOT_Seagrasses.csv"]] + theme(legend.position = "none")) /
  plot_list[["GO_results_genes_union_Aquatics_and_Seagrasses_NOT_Terrestrials.csv"]] + plot_layout(guides = 'collect') + plot_annotation(tag_levels = 'A')
patchy
cowplot::save_plot('output/GO_results_terrestrials_vs_union_aquatics_seagrasses.png', patchy, base_height=7)
```

OK I'll better make this figure for the seagrasses alone, it's quite messy this way.

# Seagrass-specific loss

Let's check which GO terms overlap between the 4 seagrasses!

How many shared lost GO-terms are there?

```{r}
a <- list(`P. australis` = results_list$GO_results_genes_lost_P_australis_not_other_seagrasses.csv$Name,
          `A. antarctica` = results_list$GO_results_genes_lost_A_antarctica_not_other_seagrasses.csv$Name,
          `Z. marina` = results_list$GO_results_genes_lost_Z_marina_not_other_seagrasses.csv$Name,
          `Z. muelleri` = results_list$GO_results_genes_lost_Z_muelleri_not_other_seagrasses.csv$Name)

a_go_ids <- list(`P. australis` = results_list$GO_results_genes_lost_P_australis_not_other_seagrasses.csv$TermID,
          `A. antarctica` = results_list$GO_results_genes_lost_A_antarctica_not_other_seagrasses.csv$TermID,
          `Z. marina` = results_list$GO_results_genes_lost_Z_marina_not_other_seagrasses.csv$TermID,
          `Z. muelleri` = results_list$GO_results_genes_lost_Z_muelleri_not_other_seagrasses.csv$TermID)
```

```{r}
plot(venn(a), 
     quantities = TRUE,
     fill = rev(wes_palette("Zissou1", 15, type = 'continuous')),
    alpha = 0.5,
     labels = list(font = 4))
```

@@

What are the shared GO-terms in seagrasses, WITHOUT the Ara loss??
```{r}
Reduce(union, a) %>% enframe() %>% writexl::write_xlsx('./output/Seagrasses_shared_lost_genes.xlsx')
```

@@

```{r}
writexl::write_xlsx(results_d, 'output/Lost_GO_terms_in_five_species.xlsx')
```

# After filtering for plant-specific GO-terms

@@

```{r}
results_d %>% filter(GOID %in% plantSpecificGO) %>% writexl::write_xlsx('output/Lost_GO_terms_in_five_species.PlantSpecific.xlsx')
```

@@

```{r}
plot(venn(newa), 
     quantities = TRUE,
     fill = rev(wes_palette("Zissou1", 15, type = 'continuous')),
    alpha = 0.5,
     labels = list(font = 4))
```
